D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-04-07 03labelled.tex
.
update the proof

\newcommand {\cD } {{\mathcal{D}}}
\newcommand {\cl } {{\mathit{cl}}}

@@

\paragraph*{2. Rewriting to normal form.}
We add three types of rules for dealing with operators on labelled terms, 1) recursively applying them to subterms, 2) pushing big operators out and 3) eliminating generalized composition and bra-ket pairs.
%Other symbols on labelled Dirac notation are also 
Take the rule for conjugate $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$ as an example of 1).
For 2), we use distributivity rules for scaling, labelled tensor and generalized composition. For example, rule
\[
% \textrm{(R-SUM-PUSHD0)}
X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n)
\]
will lift summation to the outside.
Extra rules for 3) are established including:
% The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
\begin{align*}
    % & \textrm{(R-SUM-PUSHD0)}
    % && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
    % %
    % & \textrm{(R-SUM-PUSHD1)}
    % && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
    %
    & \textrm{(R-L-SORT0)}
    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
    %
    & \textrm{(R-L-SORT1)}
    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
    %
    & \textrm{(R-L-SORT2)}
    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)
    %
\end{align*}
% These rules are added to the rewriting system in~\Cref{sec: decide} and executed together.
Assuming no variables of $\DType(s_1, s_2)$, 
repeating the application of the above rules yield the normal form of both sides of the equality--the addition of big operators: 
\begin{equation}
  \label{eqn:Labelled normal}
  \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
  + \cdots +
  \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
\end{equation}
where each sum body is a tensor of labelled basis with scalar coefficients in plain Dirac notation.
\paragraph*{3. Ordering and elimination of quantum variables.} 
We further sort tensors of labelled basis in every sum body of Eqn. (\ref{eqn:Labelled normal}) by 1. ket first and 2. the default order of variables. This yield the same shape of every subterm on both sides of the equality, e.g., 
\begin{equation}
  \label{eqn:Labelled normal1}
  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i}_p \otimes \cdots \otimes \ket{j}_q) \otimes (\bra{i'}_{p'}\otimes\cdots\otimes \bra{j'}_{q'}))
\end{equation}
and thus it is equivalent to decide the equivalence of additions of subterms of:
\begin{equation}
  \label{eqn:Labelled normal2}
  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i} \otimes \cdots \otimes \ket{j}) \cdot (\bra{i'}\otimes\cdots\otimes \bra{j'}))
\end{equation}
which do not involve any labels.

%Define $T_{s,s'}$ as $\SType$ if $s=s'=\emptyset$, $\KType(\sigma_s)$ if $s'=\emptyset$, $\BType(\sigma_{s'})$ if $s=\emptyset$ and $\OType(\sigma_s,\sigma_{s'})$ otherwise (see Sec. \ref{context: sigma s}). We claim the soundness of our normalization as:
% Then if the original equality is $\Gamma\vdash D_1 = D_2 : \cD(s,s')$, what remains to be check $\Gamma\vdash e_1 = e_2 : T_{s,s'}$.
\paragraph*{}
By routinely showing that Theorem \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules} (regard rules as equational theory), we further claim:
\begin{theorem}[Soundness and Completeness of Normalization]
  \label{thm: normalization}
  Assume $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 : \cD(s,s')$
  and no variables of $\DType(\cdot,\cdot)$ appear in $D_1,D_2$.
  Let $e_1 = e_2$ be obtained by above normalization procedure on $D_1 = D_2$. Then $\Gamma\vDash D_1 = D_2$ if and only if $\Gamma\vDash e_1 = e_2$.
  %All rewriting rules for labelled Dirac notations listed here and in Appendix are sound, i.e., for any interpretation $\sem{\cdot}$, $D\reduce D'$ implies $\sem{D} = \sem{D'}$.
\end{theorem}
% \begin{proof}
%   Notice that all rewriting rules and ordering perserve the types, so the final step eliminating labels is both sound and complete. 
%   It is routinely to check that each rewriting rules and ordering step are sound based on the semantics. This directly leads to the soundness. For completeness, we need further check if the normalization procedure always provide $e_1 = e_2$, i.e., it never gets stuck and terminates. Step 1 is trivial. For step 2, it can be further checked that 1. there is no circular rules and thus terminates, and 2. if it is not in form (\ref{eqn:Labelled normal}), there must exists one of the rules can be further applied. For step 3, ordering is always applicable. This completes the proof.
% \end{proof}

@@

% In this stage, we only need to check the scalars to decide the equivalence of labelled Dirac notation.

>>>>

@@

\paragraph*{2. Rewriting to normal form.}
We add three types of rules for dealing with operators on labelled terms, 1) recursively applying them to subterms, 2) pushing big operators out and 3) eliminating generalized composition and bra-ket pairs.
%Other symbols on labelled Dirac notation are also 
Take the rule for conjugate $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$ as an example of 1).
For 2), we use distributivity rules for scaling, labelled tensor and generalized composition. For example, rule
\[
% \textrm{(R-SUM-PUSHD0)}
X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n)
\]
will lift summation to the outside.
Extra rules for 3) are established including:
% The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
\begin{align*}
    % & \textrm{(R-SUM-PUSHD0)}
    % && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
    % %
    % & \textrm{(R-SUM-PUSHD1)}
    % && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
    %
    & \textrm{(R-L-SORT0)}
    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
    %
    & \textrm{(R-L-SORT1)}
    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
    %
    & \textrm{(R-L-SORT2)}
    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)
    %
\end{align*}
% These rules are added to the rewriting system in~\Cref{sec: decide} and executed together.
Assuming no variables of $\DType(s_1, s_2)$, 
repeating the application of the above rules yield the normal form of both sides of the equality--the addition of big operators: 
\begin{equation}
  \label{eqn:Labelled normal}
  \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
  + \cdots +
  \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
\end{equation}
where each sum body is a tensor of labelled basis with scalar coefficients in plain Dirac notation.
\paragraph*{3. Ordering and elimination of quantum variables.} 
We further sort tensors of labelled basis in every sum body of Eqn. (\ref{eqn:Labelled normal}) by 1. ket first and 2. the default order of variables. This yield the same shape of every subterm on both sides of the equality, e.g., 
\begin{equation}
  \label{eqn:Labelled normal1}
  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i}_p \otimes \cdots \otimes \ket{j}_q) \otimes (\bra{i'}_{p'}\otimes\cdots\otimes \bra{j'}_{q'}))
\end{equation}
and thus it is equivalent to prove the equivalence of additions of subterms of:
\begin{equation}
  \label{eqn:Labelled normal2}
  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i} \otimes \cdots \otimes \ket{j}) \cdot (\bra{i'}\otimes\cdots\otimes \bra{j'}))
\end{equation}
which do not involve any labels.

%Define $T_{s,s'}$ as $\SType$ if $s=s'=\emptyset$, $\KType(\sigma_s)$ if $s'=\emptyset$, $\BType(\sigma_{s'})$ if $s=\emptyset$ and $\OType(\sigma_s,\sigma_{s'})$ otherwise (see Sec. \ref{context: sigma s}). We claim the soundness of our normalization as:
% Then if the original equality is $\Gamma\vdash D_1 = D_2 : \cD(s,s')$, what remains to be check $\Gamma\vdash e_1 = e_2 : T_{s,s'}$.
\paragraph*{}
We claim that Theorem \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules} (regard rules as equational theory), as well as the relative completeness of the normalization with respect to plain Dirac notation:
\begin{theorem}[Soundness and Completeness of Normalization]
  \label{thm: normalization}
  Assume $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 : \cD(s,s')$
  and no variables of $\DType(\cdot,\cdot)$ appear in $D_1,D_2$.
  Let $e_1 = e_2$ be obtained by above normalization procedure on $D_1 = D_2$. Then $\Gamma\vDash D_1 = D_2$ if and only if $\Gamma\vDash e_1 = e_2$.
  %All rewriting rules for labelled Dirac notations listed here and in Appendix are sound, i.e., for any interpretation $\sem{\cdot}$, $D\reduce D'$ implies $\sem{D} = \sem{D'}$.
\end{theorem}
\begin{hproof}
  \lz{
  Checking the soundness of every rule is routine. 
  For relative completeness, it is sufficient to show that the normalization steps can succeed for every labelled Dirac equality. To this end, we give 
  concrete steps to rewrite every labelled Dirac notation into forms of Eqn. (1) by induction on the structure of the notation. Further noticing that reordering of labelled basis preserves the type and semantics and thus yield Eqn. (4). The detailed proof can be found in Appendix \ref{sec: rewriting rules}.
  }

  \lz{Note that since we take a special algorithm (procedure) to rewrite equations, the relative completeness is weaker than showing the confluence and termination of all these rewriting rules.}
\end{hproof}

% \begin{proof}
%   Notice that all rewriting rules and ordering perserve the types, so the final step eliminating labels is both sound and complete. 
%   It is routinely to check that each rewriting rules and ordering step are sound based on the semantics. This directly leads to the soundness. For completeness, we need further check if the normalization procedure always provide $e_1 = e_2$, i.e., it never gets stuck and terminates. Step 1 is trivial. For step 2, it can be further checked that 1. there is no circular rules and thus terminates, and 2. if it is not in form (\ref{eqn:Labelled normal}), there must exists one of the rules can be further applied. For step 3, ordering is always applicable. This completes the proof.
% \end{proof}

@@

% In this stage, we only need to check the scalars to prove the equivalence of labelled Dirac notation.
